### 2017-04-28 paper/jfp-submission/evaluation.tex
update discussion section

There are, of course, drawbacks to our approach. Four that stand out
are:
%
(1) coverage limits due to random generation;
%
(2) the inability to handle certain instances of infinite types;
%
(3) dealing with an explosion in the size of generated traces; and
%
(4) handling ad-hoc polymorphism.

@@

\paragraph{Infinite Types}
Our implementation does check for infinite types inside \forcesym, but
there are some degenerate cases where it is unable to detect
them. Consider, the following buggy @replicate@
%
\begin{code}
  let rec replicate n x =
    if n <= 0 then
      []
    else
      replicate (n-1) [x]
\end{code}
%
This code produces a nested list (with @n@ levels of nesting) containing
a single copy of @x@, instead of a list with @n@ copies of @x@. \ocaml\
detects a cyclic \hbox{@'a = 'a list@} constraint in the recursive call
and throws a type error, whereas \nanomaly\ happily % recurses @n@ times to
produces the nested list.  Strictly speaking, this function itself cannot
``go wrong'', the program would not get stuck until a \emph{client}
attempted to use the result expecting a flat list. But this is not very
satisfying as @replicate@ is clearly to blame. Furthermore, in our
experience, infinite-type errors are often difficult to %some of the more difficult ones to
debug (and to explain to novices), so better support for this scenario
would be useful.

@@

\paragraph{Ad-hoc Polymorphism}
Our approach can only support ad-hoc polymorphism (\eg\ type-classes in
\haskell\ or polymorphic comparison functions in \ocaml) in limited cases
where we have enough typing information at the call-site to resolve the
overloading. For example, consider the @n <= 0@ test in our @fac@ example.
@<=@ is polymorphic in \ocaml, but in this case we can make progress because
the literal @0@ is not. If we parameterized @fac@ by a lower bound, \eg
%
\begin{code}
  let rec fac n m =
    if n <= m then
      1
    else
      n * fac (n - 1) m
\end{code}
%
and called @fac@ with two holes, we would get stuck at the @n <= m@
test; not because of a type error, but because all we know about
@n@ and @m@ at that point is that they must have the same (unknown)
type.

This issue is uncommon in \ocaml\
(we detected it in around 5\% of our benchmarks),
but it would surely be exacerbated
by a language like \haskell, which makes heavy use of overloading. We
suspect that dynamic-symbolic execution would allow us to handle ad-hoc
polymorphism, but defer a proper treatment to future work.

>>>>

There are, of course, drawbacks to our approach. Four that stand out
are:
%
(1) coverage limits due to random generation;
%
(2) dealing with an explosion in the size of generated traces;
%
%(3) the inability to handle certain instances of infinite types; and
(3) our use of a non-parametric function type; and
%
(4) handling ad-hoc polymorphism.

@@

% \paragraph{Infinite Types}
% Our implementation does check for infinite types inside \forcesym, but
% there are some degenerate cases where it is unable to detect
% them. Consider the following buggy @replicate@.
% %
% \begin{code}
%   let rec replicate n x =
%     if n <= 0 then
%       []
%     else
%       replicate (n-1) [x]
% \end{code}
% %
% This code produces a nested list (with @n@ levels of nesting) containing
% a single copy of @x@, instead of a list with @n@ copies of @x@. \ocaml\
% detects a cyclic \hbox{@'a = 'a list@} constraint in the recursive call
% and throws a type error, whereas \nanomaly\ happily recurses @n@ times to
% produces the nested list.  Strictly speaking, this function itself cannot
% ``go wrong'', the program would not get stuck until a \emph{client}
% attempted to use the result expecting a flat list. But this is not very
% satisfying as @replicate@ is clearly to blame. Furthermore, in our
% experience, infinite-type errors are often difficult to %some of the more difficult ones to
% debug (and to explain to novices), so better support for this scenario
% would be useful.

@@

\paragraph{Non-Parametric Function Type}
As we discussed in \S~\ref{sec:how-safe} some ill-typed programs
lack a witness in our semantics due to our use of a non-parametric
type $\tfun$ for functions.
%
These programs cannot ``go wrong'', strictly speaking, but would be very
difficult to \emph{use} in practice.
%
We also note that many of these programs induce cyclic typing constraints,
causing infinite-type errors, which in our experience can be particularly
difficult to debug (and to explain to novices).

Better support for these programs would be welcome.
%
For example, we might track how the types of inputs change between
recursive calls.
%
If we cannot find a traditional witness, we could then produce a trace
expanded to show these particular steps.

\paragraph{Ad-Hoc Polymorphism}
% Our approach can only support ad-hoc polymorphism (\eg\ type-classes in
% \haskell\ or polymorphic comparison functions in \ocaml) in limited cases
% where we have enough typing information at the call-site to resolve the
% overloading. For example, consider the @n <= 0@ test in our @fac@ example.
% @<=@ is polymorphic in \ocaml, but in this case we can make progress because
% the literal @0@ is not. If we parameterized @fac@ by a lower bound, \eg
% %
% \begin{code}
%   let rec fac n m =
%     if n <= m then
%       1
%     else
%       n * fac (n - 1) m
% \end{code}
% %
% and called @fac@ with two holes, we would get stuck at the @n <= m@
% test; not because of a type error, but because all we know about
% @n@ and @m@ at that point is that they must have the same (unknown)
% type.

Also discussed in \S~\ref{sec:how-safe}, our approach can only support
ad-hoc polymorphism (\eg\ type-classes in \haskell\ or polymorphic
comparison functions in \ocaml) in limited cases where we have enough
typing information at the call-site to resolve the overloading. This
issue is uncommon in \ocaml\ (we detected it in around 5\% of our
benchmarks), but it would surely be exacerbated by a language like
\haskell, which makes heavy use of overloading. We suspect that either
dynamic-symbolic execution or speculative instantiation of holes would
allow us to handle ad-hoc polymorphism, but defer a proper treatment to
future work.

### 2017-04-28 paper/jfp-submission/safe.tex
use nate-style color boxes
update discussion section

\subsubsection{How Safe are the ``Safe'' Programs?}
\label{sec:how-safe}

An immediate question arises regarding the 13--16\% of programs for
which we could not synthesize a witness:
%
are they \emph{actually} safe (\ie is the type system being too conservative),
%
or did \toolname simply fail to find the witness?
%
% \ES{Need better organization for the groupings of programs, maybe a pie chart to guide us?}

@@

\paragraph{Overloaded Functions}
%
We found that for 5\% programs \toolname got stuck when it tried to
compare two holes.
%
\ocaml provides polymorphic equality and comparison operators,
overloading them for each type.
%
While convenient to use, they pose a challenge for \toolname's
combination of execution and inference.
%
Consider, for example
%
\begin{code}
  let bad x y =
    if x < y then
      1 + true
    else
      0
\end{code}
%
When given @bad@, \toolname will generate two fresh holes
$\nu_1[\alpha_1]$ and $\nu_2[\alpha_2]$ and proceed directly into the
@x < y@ comparison.
%
We cannot (yet) instantiate either hole because we have no constraints
on the $\alpha$s (we know they must be equal, but nothing else), and
furthermore we do not know what constraints we may encounter later on in
the program.
%
Thus, we cannot perform the comparison and proceed, and must give up our
search for a witness, even though one obviously exists, any pair of |x|
and |y| such that |x < y| is true.

Extending \toolname with support for symbolic execution would alleviate
this issue, as we could then begin symbolically executing the program
until we learn how to instantiate |x| and |y|.
%
Alternatively, we could \emph{speculatively} instantiate both |x| and
|y| with some arbitrary type, and proceed with execution until we
discover a type error.
%
This speculative instantiation is of course unsound; we would have to
take care to avoid reporting frivolous type errors that were caused by
such instantiations.
%
We would need to track which holes were instantiated speculatively in
order to distinguish type errors that would have happened regardless, as
in @bad@, from type errors that were caused by our instantiation.

Further, suppose that our speculative instantiation induces a frivolous
type error.
%
For example, suppose we are given \pagebreak
%
\begin{code}
  let bad x y =
    if x < y then
      x *. y
    else
      0.0
\end{code}
%
and choose to (speculatively) instantiate @x@ and @y@ as @int@s and proceed
down the ``true'' branch.
%
We will quickly discover this was the wrong choice, as they are immediately
narrowed to @float@s.
%
We must now backtrack and try a different instantiation, but we no
longer need to choose one at random.
%
Since our instantiation was speculative, we can still view @x@ and @y@
\emph{morally} as holes, and thus treat the @*.@ operator as a normal
narrowing point with two holes.
%
This tells us that the \emph{correct} instantiation was in fact @float@,
and we can then proceed as normal from the backtracking point with a
concrete choice of @float@s.
%
Thus, it appears that speculative instantiation of holes may be a
useful, lightweight alternative to symbolic execution for our purposes.

>>>>

\subsubsection{How safe are the ``safe'' programs?}
\label{sec:how-safe}

An immediate question arises regarding the 13--16\% of programs for
which we could not synthesize a witness:
%
are they \emph{actually} safe (\ie is the type system being too conservative),
%
or did \toolname simply fail to find a witness?
%
% \ES{Need better organization for the groupings of programs, maybe a pie chart to guide us?}

@@

\paragraph{Ad-Hoc Polymorphism}
%
We found that for 5\% programs \toolname got stuck when it tried to
compare two holes.
%
\ocaml provides polymorphic equality and comparison operators,
overloading them for each type.
%
While convenient to use, they pose a challenge for \toolname's
combination of execution and inference.
%
For example, consider the \hbox{@n <= 0@} test in our @fac@ example.
%
The @<=@ operator is polymorphic, but in this case we can make progress because
the literal @0@ is not.
%
Suppose, however, we parameterized @fac@ by a lower bound, \eg
%
\begin{code}
  let rec fac n m =
    if n <= m then
      true
    else
      n * fac (n - 1) m
\end{code}
%
When given @fac@, \toolname will generate two fresh holes
$\nu_1[\alpha_1]$ and $\nu_2[\alpha_2]$ and proceed directly into the
@n < m@ comparison.
%
We cannot (yet) instantiate either hole because we have no constraints
on the $\alpha$s (we know they must be equal, but nothing else), and
furthermore we do not know what constraints we may encounter later on in
the program.
%
Thus, we cannot perform the comparison and proceed, and must give up our
search for a witness, even though one obviously exists, any pair of |n|
and |m| such that |n <= m| is false.

Extending \toolname with support for symbolic execution would alleviate
this issue, as we could then begin symbolically executing the program
until we learn how to instantiate |n| and |m|.
%
Alternatively, we could \emph{speculatively} instantiate both |n| and
|m| with some arbitrary type, and proceed with execution until we
discover a type error.
%
This speculative instantiation is of course unsound; we would have to
take care to avoid reporting frivolous type errors that were caused by
such instantiations.
%
We would need to track which holes were instantiated speculatively in
order to distinguish type errors that would have happened regardless, as
in @fac@, from type errors that were caused by our instantiation.

Further, suppose that our speculative instantiation induces a frivolous
type error.
%
For example, suppose we are given
%
\begin{code}
  let bad x y =
    if x < y then
      x *. y
    else
      0.0
\end{code}
%
and choose to (speculatively) instantiate @x@ and @y@ as @int@s and proceed
down the ``true'' branch.
%
We will quickly discover this was the wrong choice, as they are immediately
narrowed to @float@s.
%
We must now backtrack and try a different instantiation, but we no
longer need to choose one at random.
%
Since our instantiation was speculative, we can still view @x@ and @y@
\emph{morally} as holes, and thus treat the @*.@ operator as a normal
narrowing point with two holes.
%
This tells us that the \emph{correct} instantiation was in fact @float@,
and we can then proceed as normal from the backtracking point with a
concrete choice of @float@s.
%
Thus, it appears that speculative instantiation of holes may be a
useful, lightweight alternative to symbolic execution for our purposes.
